Neural Power Units

### 2020-09-30 papers/main.tex
add authors
fix definitions

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage{neurips_2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsthm}         % definitions
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{caption}
\usepackage{mathtools} 
\usepackage{amsthm}

@@

\title{Neural Power Units}
\author{Anonymous Authors}

@@

\citet{trask_neural_2018} have demonstrated the severity of the extrapolation
problem of dense networks for even the simplest arithmetic operations, such as
summing or multiplying two numbers.  In order to increase the power of
abstraction for arithmetic tasks, they propose the \emph{Neural Arithmetic
Logic Unit} (NALU), which is capable of learning ($+,\times,\div$).  However,
the NALU cannot handle negative inputs correctly due to the logarithm in
Eq.~\ref{eq:nalu_mult}:
\begin{definition}[{\bf NALU}]
  The NALU consits of a ($+$) and a ($\times$) path that share their weights $\bm{\hat{W}}$ and $\bm{\hat{M}}$.
\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm{\hat W}& = \tanh(\bm{W}) \odot \sigma(\bm{M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm{\hat W}(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}
\end{definition}
Additionally, the logarithm destabilizes training to the extent that the chance of success
can drop below 20\% for ($+,\times$), it becomes practically impossible to
learn ($\div$) and difficult to learn from small inputs in general
\citep{madsen_measuring_2019}.
\cite{schlor_inalu_2020} provide a detailed description of the shortcomings of
the NALU, and they suggest an \emph{improved NALU} (iNALU).  The iNALU addresses
the NALU's problems through several mechanisms.  It has independent
addition and multiplication weights for Eq.~\ref{eq:nalu_add} and
Eq.~\ref{eq:nalu_mult}, clips weights and gradients to improve training
stability, regularizes the weights to push them away from zero, and, most
importantly, introduces a mechanism to recover the sign that is lost due to the
absolute value in the logarithm. Additionally, the authors propose to
reinitialize the network if its loss is not improving during training.  We
include the iNALU in one of our experiments and find that it only slightly
improves the NALU's performance (Sec.~\ref{sub:simple_arithmetic_task}) at the
cost of a significantly more complicated unit. Our NPU avoids all these mechanisms
by internally using complex arithmetics.

@@

Instead of trying to fix the NALU's convergence issues,
\cite{madsen_neural_2020} propose a new unit for ($\times$) only.
The \emph{Neural Multiplication Unit} (NMU) uses explicit multiplications and learns
to gate between identity and ($\times$) of inputs. The NMU is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{definition}[{\bf NMU \& NAU}]
  NMU and NAU are two units that can be stacked to model ($+,\times$).
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i \hat M_{ij} z_{i} + 1 - \hat M_{ij}  & \hat M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm{\hat{A}} \bm x & \hat A_{ij}=\min(\max(A_{ij}, -1), 1)
\end{align}
\end{definition}
Both NMU and NAU are regularized with $\mathcal{R} = \sum_{ij} \min(|W_{ij}|,
|1-W_{ij}|)$, and their weights are clipped, which biases them towards learning
an operation or pruning it completely.  The combination of NAU and NMU can thus
learn $(+,\times)$ for both positive and negative inputs. Training NAU and NMU
is stable and succeeds much more frequently than with the NALU, but they cannot
represent ($\div$), which we address with our NPU.

@@

With the modifications introduced above we can extend the multiplication path
of the NALU from
\begin{align}
  \label{eq:nalu_mult_2}
  \bm m = \exp \bm W(\log_{\text{real}}(|\bm x|+\epsilon))
\end{align}
to use the complex logarithm ($\log\coloneqq\log_{\text{complex}}$) and a
complex weight $\bm W$ to
\begin{align}
  \label{eq:1}
  \bm z = \exp(\bm W\log \bm x) = \exp\left((\bm \Wre + i\Wim) \log\bm x\right).
\end{align}
The complex log in Eq.~\ref{eq:1} lifts the positivity constraint on $\bm x$ resulting
in a layer that can process both positive and negative numbers correctly.  A complex
weight matrix $\bm W$ somewhere in a larger network would result in complex
gradients in other layers.  This would effectively result in doubling the
number of parameters of the whole network. As we are only interested in real
networks outputs, we can avoid this doubling by considering only the real part
of the output $\bm z$:
\begin{align}
  \real(\bm z) &= \real(\exp((\Wre + i\Wim)(\log\bm r + i\pi\bm k))) \\
    \label{eq:npumult}
    &= \exp(\Wre\log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log\bm r + \pi\Wre\bm k).
\end{align}
Above we have used Euler's formula\footnote{Euler's formula states that for any
real number $x$: $e^{ix} = \cos x + i\sin x$} and the fact that the complex
logarithm for real valued inputs is
\begin{align}
  \log x = \log r + i\theta = \log r + ik\pi,
\end{align}
where $k=0$ if $r\geq0$ and $k=1$ if $r<0$.
A diagram of the NaiveNPU is shown in Fig.~\ref{fig:npu_diagram}.
\begin{definition}[{\bf NaiveNPU}]
  The Naive Neural Power Unit with matrices $\Wre$ and
  $\Wim$ representing real and imaginary part of the complex numbers defined as
  \begin{gather}
    \label{eq:naivenpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where }\\
    \nonumber
    \bm r = |\bm x|,
    \quad
    k_i = \begin{cases}
       0 & x_i \leq 0 \\
       1 & x_i > 0
    \end{cases}
  \end{gather}
\end{definition}

@@

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs and neglecting the second one, $f(x_1,x_2) = x_1$:
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts the gradient norm of NPU and
NaiveNPU for a batch of two-dimensional inputs. One input is small and
irrelevant. Even in this simple example, gradient of the NaiveNPU is close to
zero in large parts of the parameter space This can be explained as follows.
One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  Therefore, we introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
      \hat g_i & x_i > 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Norm of the gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

@@

If an ODE model is composed of dense layers, its direct interpretation is
problematic and has to be performed retrospectively. The class of models based on
SINDy is interpretable by design, however it can only provide explanation
within a linear combination of predefined set of basis functions. Thus, it
cannot learn models with unknown fractional powers. We will demonstrate that the NPU
is capable of doing so.
An example of an ODE that contains powers is a modification of the classical
epidemiological SIR model \citep{kermack1927contribution} to fractional powers
(fSIR, \cite{taghvaei_fractional_2020}), which was shown to be a beneficial
modification for modelling the COVID-19 outbreak.  The classical SIR model is
built from three variables: $S$ (susceptible), $I$ (infectious), and $R$
(recovered/removed).  Arguably the most important part of the model is the
transmission rate $r$, which is typically taken to be proportional to the
product of $S$ and $I$.  \citet{taghvaei_fractional_2020} argue that,
especially in the initial phase of an epidemic, the boundary areas of infected
and susceptible cells scale with a fractional power, which leads to
Eq.~\ref{eq:fracsir_r}:
\begin{align}
  \label{eq:fracsir}
  \frac{dS}{dt} &= -r(t) + \eta R(t), &
  \frac{dI}{dt} &=  r(t) - \alpha I(t), &
  \frac{dR}{dt} &= \alpha I(t) - \eta R(t),\\
  \label{eq:fracsir_r}
  & & r(t) &= \beta I(t)^\gamma S(t)^\kappa,
\end{align}

>>>>

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage[final]{neurips_2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsthm}         % definitions
\usepackage{amssymb}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{caption}
\usepackage{mathtools} 
\usepackage{amsthm}

@@

\title{Neural Power Units}
\author{
    Niklas Heim, Tom\'a\v s Pevn\'y, V\'aclav \v Sm\'idl \\
    Artificial Intelligence Center\\
    Czech Technical University\\
    Prague, Czechia\\
    \texttt{heimnikl@fel.cvut.cz}
}

@@

\citet{trask_neural_2018} have demonstrated the severity of the extrapolation
problem of dense networks for even the simplest arithmetic operations, such as
summing or multiplying two numbers.  In order to increase the power of
abstraction for arithmetic tasks, they propose the \emph{Neural Arithmetic
Logic Unit} (NALU), which is capable of learning ($+,\times,\div$).  However,
the NALU cannot handle negative inputs correctly due to the logarithm in
Eq.~\ref{eq:nalu_mult}:
\begin{definition}[{\bf NALU}]
  The NALU consists of a ($+$) and a ($\times$) path that share their weights $\bm{\hat{W}}$ and $\bm{\hat{M}}$.
\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x,
                          & \bm{\hat W}& = \tanh(\bm{W}) \odot \sigma(\bm{M}), \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm{\hat W}(\log(|\bm x|+\epsilon)), & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g),
                          & \bm g& = \sigma(\bm G\bm x),
\end{align}
\end{definition}
with inputs $\bm x$ and learnt parameters $\bm W$, $\bm M$, and $\bm G$.
Additionally, the logarithm destabilizes training to the extent that the chance of success
can drop below 20\% for ($+,\times$), it becomes practically impossible to
learn ($\div$) and difficult to learn from small inputs in general
\citep{madsen_measuring_2019}.
\cite{schlor_inalu_2020} provide a detailed description of the shortcomings of
the NALU, and they suggest an \emph{improved NALU} (iNALU).  The iNALU addresses
the NALU's problems through several mechanisms.  It has independent
addition and multiplication weights for Eq.~\ref{eq:nalu_add} and
Eq.~\ref{eq:nalu_mult}, clips weights and gradients to improve training
stability, regularizes the weights to push them away from zero, and, most
importantly, introduces a mechanism to recover the sign that is lost due to the
absolute value in the logarithm. Additionally, the authors propose to
reinitialize the network if its loss is not improving during training.  We
include the iNALU in one of our experiments and find that it only slightly
improves the NALU's performance (Sec.~\ref{sub:simple_arithmetic_task}) at the
cost of a significantly more complicated unit. Our NPU avoids all these mechanisms
by internally using complex arithmetics.

@@

Instead of trying to fix the NALU's convergence issues,
\cite{madsen_neural_2020} propose a new unit for ($\times$) only.
The \emph{Neural Multiplication Unit} (NMU) uses explicit multiplications and learns
to gate between identity and ($\times$) of inputs. The NMU is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{definition}[{\bf NMU \& NAU}]
  NMU and NAU are two units that can be stacked to model ($+,\times$).
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i \hat M_{ij} x_{i} + 1 - \hat M_{ij}, & \hat M_{ij}=\min(\max(M_{ij}, 0), 1),\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm{\hat{A}} \bm x, & \hat A_{ij}=\min(\max(A_{ij}, -1), 1),
\end{align}
\end{definition}
with inputs $\bm x$, and learnt parameters $\bm M$, and $\bm A$.
Both NMU and NAU are regularized with $\mathcal{R} = \sum_{ij} \min(|W_{ij}|,
|1-W_{ij}|)$, and their weights are clipped, which biases them towards learning
an operation or pruning it completely.  The combination of NAU and NMU can thus
learn $(+,\times)$ for both positive and negative inputs. Training NAU and NMU
is stable and succeeds much more frequently than with the NALU, but they cannot
represent ($\div$), which we address with our NPU.

@@

With the modifications introduced above we can extend the multiplication path
of the NALU from
\begin{align}
  \label{eq:nalu_mult_2}
  \bm m = \exp \bm W(\log_{\text{real}}(|\bm x|+\epsilon))
\end{align}
to use the complex logarithm ($\log\coloneqq\log_{\text{complex}}$) and a
complex weight $\bm W$ to
\begin{align}
  \label{eq:1}
  \bm z = \exp(\bm W\log \bm x) = \exp\left((\bm \Wre + i\Wim) \log\bm x\right).
\end{align}
The complex log in Eq.~\ref{eq:1} lifts the positivity constraint on $\bm x$ resulting
in a layer that can process both positive and negative numbers correctly.  A complex
weight matrix $\bm W$ somewhere in a larger network would result in complex
gradients in other layers.  This would effectively result in doubling the
number of parameters of the whole network. As we are only interested in real
networks outputs, we can avoid this doubling by considering only the real part
of the output $\bm z$:
\begin{align}
  \real(\bm z) &= \real(\exp((\Wre + i\Wim)(\log\bm r + i\pi\bm k))) \\
    \label{eq:npumult}
    &= \exp(\Wre\log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log\bm r + \pi\Wre\bm k).
\end{align}
Above we have used Euler's formula\footnote{Euler's formula states that for any
real number $x$: $e^{ix} = \cos x + i\sin x$} and that the complex
logarithm for real valued inputs $x$ is
\begin{align}
  \log x = \log r + i\theta = \log r + ik\pi,
\end{align}
where $k=0$ if $x\geq0$ and $k=1$ if $x<0$.
A diagram of the NaiveNPU is shown in Fig.~\ref{fig:npu_diagram}.
\begin{definition}[{\bf NaiveNPU}]
  The Naive Neural Power Unit with matrices $\Wre$ and
  $\Wim$ representing real and imaginary part of the complex numbers is defined as
  \begin{gather}
    \label{eq:naivenpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where }\\
    \nonumber
    \bm r = |\bm x|,
    \quad
    k_i = \begin{cases}
       0 & x_i \geq 0 \\
       1 & x_i < 0
    \end{cases},
  \end{gather}
\end{definition}
with inputs $\bm x$, and learnt parameters $\Wre$, and $\Wim$.

@@

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs and neglecting the second one, $f(x_1,x_2) = x_1$:
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts the gradient norm of NPU and
NaiveNPU for a batch of two-dimensional inputs. One input is small and
irrelevant. Even in this simple example, gradient of the NaiveNPU is close to
zero in large parts of the parameter space This can be explained as follows.
One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  Therefore, we introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where } \\
    \nonumber
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \geq 0 \\
      \hat g_i & x_i < 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
with inputs $\bm x$, and learnt parameters $\Wre$, $\Wim$, $\bm g$.

\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Norm of the gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

@@

If an ODE model is composed of dense layers, its direct interpretation is
problematic and has to be performed retrospectively. The class of models based on
SINDy is interpretable by design, however it can only provide explanation
within a linear combination of predefined set of basis functions. Thus, it
cannot learn models with unknown fractional powers. We will demonstrate that the NPU
is capable of doing so.
An example of an ODE that contains powers is a modification of the classical
epidemiological SIR model \citep{kermack_contribution_1927} to fractional powers
(fSIR, \cite{taghvaei_fractional_2020}), which was shown to be a beneficial
modification for modelling the COVID-19 outbreak.  The classical SIR model is
built from three variables: $S$ (susceptible), $I$ (infectious), and $R$
(recovered/removed).  Arguably the most important part of the model is the
transmission rate $r$, which is typically taken to be proportional to the
product of $S$ and $I$.  \citet{taghvaei_fractional_2020} argue that,
especially in the initial phase of an epidemic, the boundary areas of infected
and susceptible cells scale with a fractional power, which leads to
Eq.~\ref{eq:fracsir_r}:
\begin{align}
  \label{eq:fracsir}
  \frac{dS}{dt} &= -r(t) + \eta R(t), &
  \frac{dI}{dt} &=  r(t) - \alpha I(t), &
  \frac{dR}{dt} &= \alpha I(t) - \eta R(t),\\
  \label{eq:fracsir_r}
  & & r(t) &= \beta I(t)^\gamma S(t)^\kappa,
\end{align}
